Longitudinal Principal Manifold Estimation

### 2024-06-05 paper/main.tex

Alzheimer's disease (AD) is a neurogenerative disorder affecting, among others, the structure of the brain. Longitudinal magnetic resonance imaging (MRI) data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy. Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently. As a result, there is wide variability in shape and volume estimates of brain regions of interest over time in longitudinal studies with major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials. To address this problem, we propose a longitudinal principal manifold estimation method, with the goal of recovering smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point. This approach mitigates the effects of random disturbances to the manifold between time points. We propose a novel data augmentation approach to allow the use of principal manifold estimation on self-intersecting manifolds. We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points. We show that the proposed method improves the estimation of surfaces of hippocampuses and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.

@@

@@

Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} (\cite{fefferman2016testing}). Manifold learning approaches have been applied in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}) and robotics (\cite{gao2023k}; \cite{gao2024bi}). In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis. Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures.

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data, including time in the dimension reduction process. Because of this treatment of the time dimension, it is incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.

Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}. \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, however, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability. Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

Our work was motivated by longitudinal imaging studies, where participants data are collected during several study visits to observe changes in brain structure (among other variables of interest) over time. One such study is the ADNI, where we obtained data that we analyzed to obtain the results presented in this manuscript. The data are publicly available and are hosted at \texttt{adni.loni.usc.edu}. The study has over 20 years of history with the goal of obtaining biomarkers of progression of early AD from various data sources including brain imaging (such as magnetic resonance imaging and positron emission tomography), cognitive and biological markers.

The remainder of this article is laid out as follows. In Section \ref{s:PME}, we review the principal manifold framework. In Section \ref{s:LPME}, we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show our proposed novel  algorithm for dimension reduction using the longitudinal principal manifold framework. Section \ref{s:simulations} demonstrates the performance of this approach on simulated data. In Section \ref{s:application}, we apply our proposed method to estimate smooth time dependent surfaces of the hippocampus and thalamus in individuals with AD and cognitively normal adults. The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.

\section{Principal Manifolds and Estimation Algorithm}\label{s:PME}

The framework for principal manifolds originated with the concept of principal curves (\cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud. However, as pointed out by \cite{duchamp1996extremal}, the principal curve framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection. In addition, applying the methods in high-dimensional settings is problematic. Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework. In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the $D-$dimensional data cloud, while the $d-$dimensional smooth surface of the region is the underlying manifold of interest. For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. Given a function $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$, $\nabla^{\otimes 2} f$ denotes the Hessian matrix of $f$, defined as $r \to \left( \frac{\partial ^2 f}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Most manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for two- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are then defined as follows.

@@

  Let $\boldsymbol{X}$ denote a random $D$-vector associated with the probability measure $\mathbb{P}$ such that $\boldsymbol{X}$ has compact support and finite second moments. Under some assumptions on limiting behavior and smoothness on functions $f$ and  $g$, given a $\lambda \in [0, \infty)$, and the following functionals:
  \begin{align}\label{eq:5}
&\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|\boldsymbol{X} - f(\pi_g(\boldsymbol{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \\

@@

  The principal manifold of $\boldsymbol{X}$ with the tuning parameter $\lambda$ is the manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ if

@@

where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, the set $\mathcal{D}'(\mathbb{R}^d)$ denotes the collection of generalized functions defined on $\mathbb{R}^d$ (also known as ``distributions" in the mathematical community, see Chapter 6 of \cite{rudin1991functional} for details), and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.

@@

The optimization function $\mathcal{K}_{\lambda, \mathbb{P}}$ in equation (\ref{eq:5}) consists of two elements. First is the expected distance from each point in the cloud to the fitted manifold, e.g. the distance between the fitted surface and the coordinates of the vertices on the 3-dimensional brain region of interest which we intend to minimize. The second element of the equation imposes the smoothness penalty on the estimated manifold. The choice of $\|f''(r)\|_{\mathbb{R}^{D}}^2$ provides a greater penalization of roughness than using the first derivative often implemented in this setting. To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, this term can be generalized using the total squared curvature of the manifold and defined as $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$. In this setting,  $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space. It is important to note that the  magnitude of the penalty on roughness in this framework depends on the given value of $\lambda$.

@@

The framework described above is an excellent approach for fitting a 2-dimensional principal surface that passes through the ``middle'' of a 3-dimensional cloud of vertices. In this section, we introduce an approach to extending this framework to longitudinal clouds of vertices. We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies. We consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e. $D=3$ to $(r_1, r_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest. Now consider that the data are observed longitudinally during several visits by a study participant. Hence, the observed data are the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of $j$th study visit. Figure \ref{fig:lpme_step1} displays a simplified example of this setting using simulated data where $D = 2$. The goal is to fit a surface through the data that depends on time and is not affected by the random noise added at each time point and in space.

Our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth. It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, e.g. modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc. The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method in Section \ref{s:PME} then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time. In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(y_1,y_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$. In order to address the limitations of existing approaches for the setting described above, we propose the longitudinal principal manifold estimation (LPME) framework to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.

To give a more general representation of this problem, let $\left\{\bm{x}_i\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points. As in the setting described above, at each time point $t$, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise. However, in this updated case, the function determining the manifold may change over time.

\subsection{The Longitudinal Principal Manifold Framework}

Suppose the $d$-dimensional manifold of interest depends on time $t$, which can be represented by a function $F(r,t)$ defined on the ``spacetime" $\mathbb{R}^d\times\mathbb{R}$, where $\mathbb{R}^d$ and $\mathbb{R}$ correspond to ``space" and ``time," respectively. For each fixed time point $t$, the function $f_t(r):=F(r,t)$ of $r$ is the manifold of interest, e.g. the surface of the 3D brain region, at time $t$.
\begin{definition}
  \label{def:lpme} Given $\boldsymbol{X} = \{ \boldsymbol{X}_t \}_{t=1}^T$, denoting the collection of data clouds observed at all time points, where $\boldsymbol{X}_t$ is the random  $D$-vector observed at time $t$, the manifold $M_F^{d+1}$, which is representative of $\boldsymbol{X}$, and tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, let the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ be defined as follows.
\begin{align}\label{eq:newKappa}
  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\boldsymbol{X}_t - f_t\left(\pi_{f_t}(\boldsymbol{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt, \nonumber
\end{align}
where $F \in C(\mathbb{R}^{d}\times\mathbb{R} \to \mathbb{R}^{D})$, such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(r,t)\|_{\mathbb{R}^{D}} = \infty$ and the elements of the function $F$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$. Then the manifold $M_{F^{*}}^{d+1}$ is the longitudinal principal manifold of $\boldsymbol{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
\end{definition}

By minimizing the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), our goal is to minimize the distance between each data point and its projection on the fitted manifold at each time point $t$, penalize the roughness of the fitted manifold at each time point $t$ depending on the value of the tuning parameter $\lambda_t$, and impose smoothness of the function $F$ over time regulated by the tuning parameter $\gamma$. These goals correspond to minimizing the first, second, and third additives of the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), respectively. Our proposed framework is based on considering smoothing at a fixed time point and smoothing over time as different problems needing differing solutions. Hence, this approach is different from a thin plate spline or other dimension reduction methods, which would smooth over both the manifold and time jointly or reduce the dimension of space and time together. Using separate smoothing parameters provides additional flexibility in situations where a manifold with a high level of spatial roughness shows minimal changes over time, or vice versa. Generally, our proposed framework can be used in applications where a high-dimensional subset of variables of interest set with low-dimensional representations of interest, while the resulting low-dimensional representations may be considered smooth over the other set of variables.

>>>>

Alzheimer's disease is a neurogenerative disorder affecting, among others, the structure of the brain. Longitudinal magnetic resonance imaging data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy. Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently. As a result, there is wide variability in shape and volume estimates of brain regions of interest over time in longitudinal studies with major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials. To address this problem, we propose a longitudinal principal manifold estimation method, with the goal of recovering smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point. This approach mitigates the effects of random disturbances to the manifold between time points. In addition, we propose a novel data augmentation approach to allow the use of principal manifold estimation on self-intersecting manifolds. We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points. We show that the proposed method improves the estimation of surfaces of hippocampuses and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.

@@

\subsection{A Motivating Question from the Medical Imaging}

@@

\subsection{A Brief Overview of Existing Approaches and Our Contributions}

Manifold learning refers to a set of approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis"---\textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} (\cite{fefferman2016testing}). Manifold learning approaches have been applied in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}), single-cell biology (\cite{ding2023learning}), and robotics (\cite{gao2023k}; \cite{gao2024bi}). In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis. Estimating time-dependent low-dimensional representations of these shapes is essential for investigating changes of these low-dimensional structures.

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data, including time in the dimension reduction process. Because of this treatment of the time dimension, it is incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. Unlike most commonly used manifold learning methods, this approach results in an explicit functional representation of the manifold estimate. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.

Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis method proposed by \cite{kinsonLongitudinalPrincipalComponent2020} and longitudinal functional PCA proposed by \cite{greven2011longitudinal}. \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, however, parameterizations for baseline images and longitudinal image differences were estimated separately while we seek to parameterize them simultaneously. \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability.

Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time. This work was motivated by longitudinal imaging studies, where participants data are collected during several study visits to observe changes in brain structure (among other variables of interest) over time. One such study is the ADNI, where we obtained data that we analyzed to obtain the results presented in this manuscript. The data are publicly available and are hosted at \href{https://adni.loni.usc.edu/}{\texttt{https://adni.loni.usc.edu/}}. The study has over 20 years of history with the goal of obtaining biomarkers of progression of early AD from various data sources including brain imaging (such as magnetic resonance imaging and positron emission tomography), cognitive and biological markers.

The remainder of this article is laid out as follows. In Section \ref{s:PME}, we review the principal manifold framework proposed by \cite{mengPrincipalManifoldEstimation2021}. In Section \ref{s:LPME}, we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show our proposed novel  algorithm for dimension reduction using the longitudinal principal manifold framework. Section \ref{s:simulations} demonstrates the performance of this approach on simulated data. In Section \ref{s:application}, we apply our proposed method to estimate smooth time dependent surfaces of the hippocampus and thalamus in individuals with AD and cognitively normal adults. The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.

\section{Principal Manifold Estimation Algorithm}\label{s:PME}

The framework for principal manifolds originated with the concept of principal curves (\cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud. However, as pointed out by \cite{duchamp1996extremal}, the principal curve framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection and problems with applications in high-dimensional settings. Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework. In order to present the longitudinal PME approach, we start by introducing notation used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the 3-dimensional data cloud (i.e., $D=3$), while the 2-dimensional smooth surface of the region is the underlying manifold of interest (i.e., $d=2$). For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers on continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. Given a function $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$, $\nabla^{\otimes 2} f$ denotes the Hessian matrix of $f$, defined as $r \to \left( \frac{\partial ^2 f}{\partial r_i \partial r_j}(r) \right)_{1 \leq i, j \leq d}$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Many manifold estimation methods are based on defining a projection index, denoted by $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for two- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are then defined as follows.

@@

  Let $\mathbf{X}$ denote a random $D$-vector associated with the probability measure $\mathbb{P}$ such that $\mathbf{X}$ has compact support and finite second moments. Under some assumptions on limiting behavior and smoothness on functions $f$ and  $g$, given a $\lambda \in [0, \infty)$, and the following functionals:
  \begin{align}\label{eq:pme_kappa}
&\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|\mathbf{X} - f(\pi_g(\mathbf{X}))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2, \\

@@

  The principal manifold of $\mathbf{X}$ with the tuning parameter $\lambda$ is the manifold $M_{f^{*}}^{d}$ determined by $f^{*}$ if

@@

where $\mathcal{F}(\mathbb{P})$ is the collection of functions $f \in \mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$ such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty}\|f(r)\|_{\mathbb{R}^{D}} = \infty$, the components of the function $f$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}) = \left\{u : \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d})\right\}$, and the projection function of $f$ satisfies the following equality $\sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1$.

@@

The functional $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ in equation (\ref{eq:pme_kappa}) consists of two elements. The first element is the expected squared distance from data points $\mathbf{X}$ in the cloud to their projections $f(\pi_f(\mathbf{X}))$ on the manifold $M_f^d$. The second element $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$ imposes the smoothness/curvature penalty on the estimated manifold. The coefficient $\lambda$ is a tuning parameter. In this setting,  $\pi_{f_{\lambda}^{*}}(\mathbf{X})$ maps the $D$-vector $\mathbf{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space. Thus, finding the function that minimizes $\mathcal{K}_{\lambda, \mathbb{P}}(f)$ takes the form of a penalized regression problem. \cite{mengPrincipalManifoldEstimation2021} show that this function takes spline form $$f_{(n+1), l}(r) = \sum_{j=1}^N s_{j, l} \times \eta_{4-d}\left(r - \pi_{f_{(n)}}(\mu_{j, N})\right) + \sum_{k=1}^{d + 1}\alpha_{k, l} \times p_k(r), \ l = 1, 2, \dots, D,$$
under the constraint $\sum_{j=1}^{N}s_{j, l} \times p_{k}\left(\pi_{f_{(n)}}(\mu_{j, N})\right) = 0$ for all $k = 1, 2, \dots, d + 1$ and $l = 1, 2, \dots, D$, where $\mu_{j, N}, j = 1, \dots, N$ denote points that summarize the data cloud $\bf{X}$. This spline function is specified by coefficients $s_{j, l}$ and $\alpha_{k, l}$ for $j = 1, \dots, N$, $k = 1, \dots, d+1$, and $l = 1, \dots, D$. These coefficients will be used to smooth over principal manifold estimates obtained at several time points in Section \ref{s:LPME}.

@@

The PME framework described in Section \ref{s:PME} is an excellent approach for fitting a 2-dimensional surface that passes through the ``middle'' of a 3-dimensional cloud of vertices. In this section, we introduce an approach to extending the PME framework to longitudinal clouds of vertices. We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies. We consider that 3-dimensional data clouds are observed longitudinally during several visits by a study participant. Hence, the observed data $(x_1, x_2, x_3,t_j)$ are 4-dimensional, where $t_j$ denotes the time of $j$th study visit. Figure \ref{fig:lpme_step1} displays a simplified example of this setting using simulated data where $D = 2$. The goal is to fit a surface through the data that depends on time and is not affected by the random noise added at each time point and in space.

Our goal is estimation of a smooth surface for the brain region of interest at each time point such that the changes of the surface over time are smooth. It is important for the resulting manifolds to be estimated at each time point in order to enable further inference, e.g., modeling trajectories of change in region volumes over time, identifying regions that are most affected in terms of atrophy, etc. The key reasons why the existing principal manifold estimation methods cannot be implemented in this setting are 1) if we implement the data reduction method in Section \ref{s:PME} then the resulting principal manifold will not be a function of time $t_j$, which is necessary for the resulting segmentations to be useful in practice, 2) if the manifold is estimated separately at each time point, the results may not be smooth over time. In other words, the dimension reduction should be implemented to obtain the 2-dimensional coordinates $(r_1,r_2,t_j)$ while ensuring that the resulting manifold is also a function of $t_j$. In order to address the limitations of existing approaches for the setting described above, we propose the longitudinal principal manifold estimation (LPME) framework to estimate $d$-dimensional representations of $D$-dimensional objects observed over time.

This problem can be posed for positive integer values of $D$ and $d$ with $d < D$. Let $\left\{x_{it}\right\}_{i=1, t=1}^{I_t, T}$ represent the $I = \sum_{t=1}^{T}I_t$ observations in $\mathbb{R}^D$ for each individual, where $I_t$ denotes the number of observations available at each time point, with $T$ total time points. As in the setting described above, at each time point $t$, these observations lie along a $d$-dimensional manifold and are corrupted by $D$-dimensional noise. However, in this updated case, the function determining the manifold may change over time.

\subsection{The Longitudinal Principal Manifold Framework}\label{section: The Longitudinal Principal Manifold Framework}

Suppose the $d$-dimensional manifold of interest depends on time $t$, which can be represented by a function $F(r,t)$ defined on the ``spacetime" $\mathbb{R}^d\times\mathbb{R}$, where $\mathbb{R}^d$ and $\mathbb{R}$ correspond to ``space" and ``time," respectively. For each fixed time point $t$, the function $f_t(r):=F(r,t)$ of $r$ is the manifold of interest, e.g. the surface of the 3D brain region, at time $t$.
\begin{definition}
  \label{def:lpme} Given $\mathbf{X} = \{ \mathbf{X}_t \}_{t=1}^T$, denoting the collection of data clouds observed at all time points, where $\mathbf{X}_t$ is the random  $D$-vector observed at time $t$, the manifold $M_F^{d+1}$, which is representative of $\mathbf{X}$, and tuning parameters $\lambda=\{\lambda_t\}_{t\in\mathbb{R}}$ and $\gamma$, let the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$ be defined as follows.
\begin{align}\label{eq:newKappa}
  \mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F) &= \int_\mathbb{R} \mathbb{E}\left\|\mathbf{X}_t - f_t\left(\pi_{f_t}(\mathbf{X}_t)\right)\right\|_{\mathbb{R}^{D}}^2 \, dt + \int_\mathbb{R} \lambda_t \cdot\|\nabla^{\otimes 2}f_t\|_{L^2(\mathbb{R}^{d})}^2 \, dt + \gamma\cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt \\
  &= \int_{\mathbb{R}}\mathcal{K}_{\lambda_t, \mathbb{P}}(f_t) \, dt + \gamma \cdot \int_{\mathbb{R}}\left\|\frac{\partial^2}{\partial t^2}f_t\right\|_{L^2(\mathbb{R}^d)}^2 \, dt, \nonumber
\end{align}
where $F \in C(\mathbb{R}^{d}\times\mathbb{R} \to \mathbb{R}^{D}\times\mathbb{R})$, such that $\lim_{\|r\|_{\mathbb{R}^{d}} \to \infty, t \to \infty}\|F(r,t)\|_{\mathbb{R}^{D}\times\mathbb{R}} = \infty$ and the coordinates of the function $F$ are in $\nabla^{-\otimes 2}L^2(\mathbb{R}^{d}\times\mathbb{R}) = \left\{u \in \mathcal{D}'(\mathbb{R}^{d}): \|\nabla^{\otimes 2} u\|_{\mathbb{R}^{d \times d}} \in L^2(\mathbb{R}^{d}\times\mathbb{R})\right\}$. Then the manifold $M_{F^{*}}^{d+1}$ is the longitudinal principal manifold of $\mathbf{X}$ if $F_{\lambda}^{*} = \argmin_{F}\mathcal{K}_{\lambda, \gamma, \mathbb{P}}(F)$.
\end{definition}

By minimizing the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), our goal is to minimize the distance between each data point and its projection on the fitted manifold at each time point $t$, penalize the roughness of the fitted manifold at each time point $t$ depending on the value of the tuning parameter $\lambda_t$, and impose smoothness of the function $F$ over time regularized by the tuning parameter $\gamma$. These goals correspond to minimizing the first, second, and third additives of the functional $\mathcal{K}_{\lambda, \gamma, \mathbb{P}}$ in (\ref{eq:newKappa}), respectively. Our proposed framework is based on considering smoothing at a fixed time point and smoothing over time as different problems needing differing solutions. Hence, this approach is different from a thin plate spline or other dimension reduction methods, which would smooth over both the manifold and time jointly or reduce the dimension of space and time together. Using separate smoothing parameters provides additional flexibility in situations where a manifold with a high level of spatial roughness shows minimal changes over time, or vice versa. Generally, our proposed framework can be used in applications where a high-dimensional subset of variables of interest set with low-dimensional representations of interest, while the resulting low-dimensional representations may be considered smooth over the other set of variables.
